---
title: "DEM 7283 - Splines and Generalized Additive Models"
author: "Corey Sparks, PhD"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   html_document:
    df_print: paged
    fig_height: 7
    fig_width: 7
    toc: yes
    toc_float: yes
    code_download: true
---

In this example, we will describe how to use splines to model nonlinearities in predictor variables and the **_generalized additive model_**. 


## Splines
Splines are nothing more than a set of linear or non-linear functions that are tied together to construct a model for data. 

### Piece-wise splines
In the simplest example, consider a regression function whose form changes at some point in the data. So if we see data like this:


```{r}
b0<-1; b1=.5;b2=35
x<-seq(1:100)
y1<-rnorm(70,b0+b1*x[1:70], 1) 
y2<-rnorm(30, b2, 1)

y<-c(y1, y2)
plot(y~x, main="linear fit to nonlinear data")
```

we might say, yes, I can fit a linear model to that:

```{r}
plot(y~x, main="linear fit to nonlinear data")
lin<-lm(y~x)
summary(lin)
lines(fitted(lin)~x,col=3)
plot(lin)
```

but we would be sorely mistaken, as this model clearly doesn't fit the data.

Instead, we can model the change in the underlying model structure by using a series of **_knots_** in the function, or points along the data in which the form of the function changes. 

Mathematically, knots can be written as:

$$Y(x) = \begin{Bmatrix}
F_1(x) \text {  for  } x\in [x_1, x_2]\\ 
F_2(x) \text {  for  } x\in [x_2, x_3]\\ 
\cdots \\ 
F_{k}(x) \text {  for  } x\in [x_{k-1}, x_k]\\

\end{Bmatrix}$$

Where each of the $F_k (x)$ functions imply a different form in the interval between $x\in [x_{k-1}, x_k]$, where the $k$ breaks are at a given knot in the data. 

This would generate a model that looks like this:

```{r}

sp<-(lm(y~x*I(x<71)+x*I(x>71)))
summary(sp)
plot(y~x,main="Piece-wise linear fit to nonlinear data, knot at x=70")
lines(fitted(sp)~x,col=2)
plot(sp)
```


Of course, this can be much more complicated, with multiple knots in the data, to model multiple changes in the underlying data generating function

```{r}
n <- 100 # number of data points
t <-  seq(from=0, to = 2*pi, length.out=100)
a <- 3
b <- 2
c.norm <- rnorm(n, 0, .1)
amp <- 1

# generate data and calculate "y"
set.seed(1)
y2 <- a*sin(b*t)+c.norm*amp # Gaussian/normal error

# plot results
plot(t, y2, t="l", ylim=range(y2)*c(1,1.2), main="Sine curve with gaussian noise")

```

```{r}
plot(t, y2, t="l", ylim=range(y2)*c(1,1.2), main="Terrible linear model")

abline(lm(y2~t), col=3)
plot(lm(y2~t))
```

Splines are often characterized by the number of knots used to build them

```{r}
library(splines)
library(mgcv)
plot(t, y2, t="l", ylim=range(y2)*c(1,1.2), main="Increasingly complex splines")
fit<-spline(y2,t)
lines(spline(t,y2, n = 4),col="darkgreen",lwd=2,type="l")
lines(spline(t,y2, n = 10),col="red",lwd=2,type="l")
lines(spline(t,y2, n = 20),col="blue",lwd=2,type="l")
lines(spline(t,y2, n = 30),col="orange",lwd=2,type="l")
```


## Cubic splines

The splines above were piecewise linear, we can also fit piecewise cubic splines, where at each knot a cubic polynomial is used. In the plot below, we can see that by increasing the number of knots, we better approximate the nonlinearities in the data.


```{r}

plot(t, y2, t="l", ylim=range(y2)*c(1,1.2), main="Increasingly complex splines")

fit<-smooth.spline(t, y2, df=2)
fit5<-smooth.spline(t, y2, df=5)
fit10<-smooth.spline(t, y2, df=10)
lines(fit, col="red")
lines(fit5, col="green")
lines(fit10, col="blue")

```


## Regression splines
The **_generalized additive model_** (GAM) is a modeling framework that forms a linear predictor for a regression function through the combination of both smooth and linear terms. 

The GAM model forms the linear predctor of a GLM as:

$$E(y)= \beta_0 + f(x_1) + \beta_1 x_2$$

where the $f(x_1)$ term is a regression spline of one of the variables. The models can be a mixture of linear and smooth terms. 


```{r}
library(mgcv) #one library for GAMs
library(ggplot2)
library(dplyr)

dat<-data.frame(y = y2, x = t)

gamfit<-gam(y ~ s(x), data=dat)
summary(gamfit)
plot(gamfit)


dat%>%
  ggplot(aes(x=x, y=y))+
  geom_point()+
  geom_smooth(method = "gam", formula = y ~ s(x))
```



### BRFSS DATA
```{r}
load(url("https://github.com/coreysparks/data/blob/master/brfss_2017.Rdata?raw=true"))

set.seed(1234)
#samps<-sample(1:nrow(brfss_17), size = 40000, replace=F)
#brfss_17<-brfss_17[samps,]


```

### Recode variables
```{r}
library(car)
library(dplyr)

#nice MSA name
brfss_17$mmsa_name<-substr(brfss_17$mmsaname, 1,nchar(brfss_17$mmsaname)-31)

#sex
brfss_17$male<-ifelse(brfss_17$sex==1, 1, 0)

#BMI
brfss_17$bmi<-ifelse(is.na(brfss_17$bmi5)==T, NA, brfss_17$bmi5/100)

#Healthy days
brfss_17$healthdays<-Recode(brfss_17$physhlth, recodes = "88=0; 77=NA; 99=NA")

#Healthy mental health days
brfss_17$healthmdays<-Recode(brfss_17$menthlth, recodes = "88=0; 77=NA; 99=NA")

brfss_17$badhealth<-Recode(brfss_17$genhlth, recodes="4:5=1; 1:3=0; else=NA")
#race/ethnicity
brfss_17$black<-Recode(brfss_17$racegr3, recodes="2=1; 9=NA; else=0")
brfss_17$white<-Recode(brfss_17$racegr3, recodes="1=1; 9=NA; else=0")
brfss_17$other<-Recode(brfss_17$racegr3, recodes="3:4=1; 9=NA; else=0")
brfss_17$hispanic<-Recode(brfss_17$racegr3, recodes="5=1; 9=NA; else=0")

brfss_17$race_eth<-Recode(brfss_17$racegr3, 
recodes="1='nhwhite'; 2='nh black'; 3='nh other';4='nh multirace'; 5='hispanic'; else=NA",
as.factor = T)
brfss_17$race_eth<-relevel(brfss_17$race_eth, ref = "nhwhite")

#insurance
brfss_17$ins<-Recode(brfss_17$hlthpln1, recodes ="7:9=NA; 1=1;2=0")

#income grouping
brfss_17$inc<-Recode(brfss_17$incomg, recodes = "9= NA;1='1_lt15k'; 2='2_15-25k';3='3_25-35k';4='4_35-50k';5='5_50kplus'", as.factor = T)
brfss_17$inc<-as.ordered(brfss_17$inc)
#education level
brfss_17$educ<-Recode(brfss_17$educa,
recodes="1:2='0Prim'; 3='1somehs'; 4='2hsgrad'; 5='3somecol'; 6='4colgrad';9=NA",
as.factor=T)
brfss_17$educ<-relevel(brfss_17$educ, ref='2hsgrad')

#employment
brfss_17$employ<-Recode(brfss_17$employ1,
recodes="1:2='employloyed'; 2:6='nilf'; 7='retired'; 8='unable'; else=NA",
as.factor=T)
brfss_17$employ<-relevel(brfss_17$employ, ref='employloyed')

#marital status
brfss_17$marst<-Recode(brfss_17$marital,
recodes="1='married'; 2='divorced'; 3='widowed'; 4='separated'; 5='nm';6='cohab'; else=NA",
as.factor=T)
brfss_17$marst<-relevel(brfss_17$marst, ref='married')

#Age cut into intervals
brfss_17$agec<-cut(brfss_17$age80, breaks=c(0,24,39,59,79,99))

#BMI, in the brfss_17a the bmi variable has 2 implied decimal places,
#so we must divide by 100 to get real bmi's

brfss_17$bmi<-brfss_17$bmi5/100

#smoking currently
brfss_17$smoke<-Recode(brfss_17$smoker3, 
recodes="1:2=1; 3:4=0; else=NA")
#brfss_17$smoke<-relevel(brfss_17$smoke, ref = "NeverSmoked")

brfss_17$obese<-ifelse(is.na(brfss_17$bmi)==T, NA, 
                       ifelse(brfss_17$bmi>30,1,0))

```

```{r}
brfss_17<-brfss_17%>%
  select(  ststr, agec,bmi,healthmdays, age80, educ, black, hispanic, other,smoke,obese, badhealth, male, mmsawt, mmsa_name )%>%
  filter(complete.cases(.))

```


Here is a GAM fit to the BRFSS, using smooth terms for age, bmi and number of poor mental health days

```{r, warning=FALSE}

brfgam<-gam(badhealth ~ s(age80)+ s(bmi)+ s(healthmdays)+ educ+ male+ black + hispanic+ other,
            data=brfss_17,
            weights = mmsawt/mean(mmsawt, na.rm=T),
            family=binomial)

summary(brfgam)
plot(brfgam)
```

Versus the linear terms
```{r, warning=FALSE}
brfgam_l<-gam(badhealth~age80+ bmi+ healthmdays+ educ+ male+ black+ hispanic+ other,
              data=brfss_17,
              weights = mmsawt/mean(mmsawt, na.rm=T),
              family=binomial)

summary(brfgam_l)
```


### LRT for the difference in model fits

```{r}
anova( brfgam_l,
       brfgam,
       test="Chisq")
```

### Using smooth terms in other models
The regression splines can also be used in any generalized linear model using functions from the `splines` library.

The `bs()` function is a B-spline basis, and the `ns()` is the natural spline function. The natural splines often behave better at the boundaries of the data. 

Here is an example using bmi as a continuous outcome`glm()`

```{r}
library(survey)
des<-svydesign(ids = ~1,
               stata=~ststr,
               weights = ~ mmsawt,
               data=brfss_17)

ex1<-svyglm(bmi ~ bs(age80)+ male+ black+ hispanic+ other,
    family=gaussian,
    des)

summary(ex1)
```

```{r}
newdat<-expand.grid(age80=seq(20,80,5),
                    male = c(0,1), 
                    black = mean(brfss_17$black),
                    hispanic = mean(brfss_17$hispanic),
                    other = mean(brfss_17$other))

newdat$pred<-predict(ex1, newdata=newdat, type="response")

newdat%>%
  ggplot()+
  geom_line(aes(x=age80,
                y=pred,
                color=factor(male) ))
```

```{r}
ex1<-svyglm(badhealth ~ bs(age80)+ male+ black+ hispanic+ other,
    family=binomial,
    des)
summary(ex1)
```

```{r}
newdat<-expand.grid(age80=seq(20,80,5),
                    male = c(0,1), 
                    black = mean(brfss_17$black),
                    hispanic = mean(brfss_17$hispanic),
                    other = mean(brfss_17$other))

newdat$pred<-predict(ex1, newdata=newdat, type="response")

newdat%>%
  ggplot()+
  geom_line(aes(x=age80,
                y=pred,
                color=factor(male) ))

```

