---
title: "DEM 7283 - Multiple Imputation & Missing Data"
author: "Corey Sparks, PhD"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   html_document:
    df_print: paged
    fig_height: 7
    fig_width: 7
    toc: yes
    toc_float: yes
---

This example will illustrate typical aspects of dealing with missing data. Topics will include: Mean imputation, modal imputation for categorical data, and multiple imputation of complex patterns of missing data.

For this example I am using 2016 CDC Behavioral Risk Factor Surveillance System (BRFSS) SMART county data. [Link](https://www.cdc.gov/brfss/smart/smart_2016.html)

## Missing data
- Every time we ask a question on a survey, there are a variety of reasons why we may fail to get a complete response

- This is a cause for concern for those using survey data


###  Reasons for missing data
Total nonresponse

-	This is where a respondent completely refuses to participate in the survey, or some other barrier prevents them from participating (not at home, ill, disability)

-	This is usually accounted for by survey weights

- **Noncoverage**

-	When an individual is not included in the survey's sampling frame

-	Usually adjusted for in the weights

- **Item nonresponse**
-	When a respondent fails to provide an acceptable response to a survey question

-	Could be refusal

-	Could be "I don't know"

-	Gould give an inconsistent response

-	The interviewer could fail to ask the question


- **Partial nonresponse**
-	Somewhere between total nonresponse and item nonresponse

-	Respondent could end the interview

-	e.g. Incomplete responses for a wave in a panel study

-	Can be dealt with by using weights, which eliminates respondent with partial responses

-	Can be dealt with by imputation, where the value is filled in by data editors

### Types of missing data

- **Missing completely at random (MCAR)**
-	Missing values do not depend on any characteristic of an individual

- Missing at random (MAR)

-	Missing values do not relate to the item of interest itself, but it could be related to another characteristic of the respondent

-	These two are said to be "ignorable"

- **Missing Not at random (MNAR)**
-	Think of a question on satisfaction, those that are extremely dissatisfied may be less likely to respond, so the results may be biased because they don't include those folks

-	This is "non-ignorable"


### How can we tell is a variable is MCAR?
- One way to estimate if an element is MCAR or MAR is to form a missing data variable (1=missing, 0=nonmissing) and estimate a logistic regression for that variable using key characteristics of a respondent

- If all these characteristics show insignificant effects, we can pretty much assume the element is MCAR/MAR

### What does the computer do?

- Typically the computer will do one of two things

-  Delete the entire case if any variables in the equation are missing, this is called **_listwise deletion_**

- delete the case for a particular comparison, this is called **_pairwise deletion_**

- Both of these lead to fewer cases for entire models or for particular tests 

### How can we impute a missing value?
- There are easy ways and hard ways, but the answer is yes.

- Easy ways == bad ways

-	Mean substitution

-	Plugs in the value for the average response for a missing value for all individuals

-	If a large % of the respondents are missing the values (say >20%) the mean could be not estimated very well

-	The variance of the variable will be driven down, if everyone who is missing is given the mean 

-	Can lead to lower effect sizes/betas in the models using the data

### Regression imputation

- Multiple Imputation will use other characteristics of individuals with complete observations to predict the missing value for the missing person

- This would use the observed characteristics of the respondent to predict their missing value

-	Think regression!

-	Income is a good variable like this

-	People don't like to report their income, but they may be willing to report their education, and other characteristics

-	We can use those to predict their income

-	This is sensitive to assumptions of the regression model being used!

### Multiple Imputation

- This works like the regression imputation method, but applies it recursively to all patterns of missing data

- Typically this will be done several times, usually 5 is the rule most programs use, and the differences in the imputations are compared across runs

- i.e. if we impute the values then run a regression for the outcome, how sensitive are the results of the model to the imputation

- There are various ways to do this, and it depends on the scale  of the variable

###  Hot deck imputation

-	Find a set of variables that are associated with your outcome

-	If there are cases similar to yours (i.e. they are similar on the non-missing variables), you can plug in their variable of interest for the missing case

-	This is what the census did for years

### Predictive mean matching

-	Regress Y on the observed X's for the incomplete cases

-	Form fitted or "predicted" values

-	Replace the missing values with a fitted value from the regression from non-missing observations who are "close" to the missing cases

-	i.e. replace the missing values with fitted values from similar individuals

```{r, message=FALSE}
library(car)
library(mice)
library(ggplot2)
library(dplyr)
```

```{r}
load(url("https://github.com/coreysparks/data/blob/master/brfss_2017.Rdata?raw=true"))

set.seed(1234)
samp<-sample(1:dim(brfss_17)[1], size = 25000) #smaller sample for brevity
brfss_17<-brfss_17[samp,]
```


### Recodes
```{r}
#Healthy days
brfss_17$healthdays<-Recode(brfss_17$physhlth, recodes = "88=0; 77=NA; 99=NA")

#Healthy mental health days
brfss_17$healthmdays<-Recode(brfss_17$menthlth, recodes = "88=0; 77=NA; 99=NA")

brfss_17$badhealth<-Recode(brfss_17$genhlth, recodes="4:5=1; 1:3=0; else=NA")
#race/ethnicity
brfss_17$black<-Recode(brfss_17$racegr3, recodes="2=1; 9=NA; else=0")
brfss_17$white<-Recode(brfss_17$racegr3, recodes="1=1; 9=NA; else=0")
brfss_17$other<-Recode(brfss_17$racegr3, recodes="3:4=1; 9=NA; else=0")
brfss_17$hispanic<-Recode(brfss_17$racegr3, recodes="5=1; 9=NA; else=0")

brfss_17$race_eth<-Recode(brfss_17$racegr3, 
recodes="1='nhwhite'; 2='nh black'; 3='nh other';4='nh multirace'; 5='hispanic'; else=NA",
as.factor = T)
brfss_17$race_eth<-relevel(brfss_17$race_eth, ref = "nhwhite")

#insurance
brfss_17$ins<-Recode(brfss_17$hlthpln1, recodes ="7:9=NA; 1=1;2=0")

#income grouping
brfss_17$inc<-Recode(brfss_17$incomg, recodes = "9= NA;1='1_lt15k'; 2='2_15-25k';3='3_25-35k';4='4_35-50k';5='5_50kplus'", as.factor = T)
#brfss_17$inc<-as.ordered(brfss_17$inc)
#education level
brfss_17$educ<-Recode(brfss_17$educa,
recodes="1:2='0Prim'; 3='1somehs'; 4='2hsgrad'; 5='3somecol'; 6='4colgrad';9=NA",
as.factor=T)
brfss_17$educ<-relevel(brfss_17$educ, ref='2hsgrad')

#employment
brfss_17$employ<-Recode(brfss_17$employ1,
recodes="1:2='employloyed'; 2:6='nilf'; 7='retired'; 8='unable'; else=NA",
as.factor=T)
brfss_17$employ<-relevel(brfss_17$employ, ref='employloyed')

#marital status
brfss_17$marst<-Recode(brfss_17$marital,
recodes="1='married'; 2='divorced'; 3='widowed'; 4='separated'; 5='nm';6='cohab'; else=NA",
as.factor=T)
brfss_17$marst<-relevel(brfss_17$marst, ref='married')

#Age cut into intervals
brfss_17$agec<-cut(brfss_17$age80, breaks=c(0,24,39,59,79,99))

#BMI, in the brfss_17a the bmi variable has 2 implied decimal places,
#so we must divide by 100 to get real bmi's

brfss_17$bmi<-brfss_17$bmi5/100

#smoking currently
brfss_17$smoke<-Recode(brfss_17$smoker3, 
recodes="1:2=1; 3:4=0; else=NA")
#brfss_17$smoke<-relevel(brfss_17$smoke, ref = "NeverSmoked")


```

Now, we can get a general idea of the missingness of these variables by just using `summary(brfss_17)`

```{r}
summary(brfss_17[, c("ins", "smoke",  "bmi", "badhealth", "race_eth",  "educ", "employ", "marst", "inc")])
```

Which shows that, among these recoded variables, `inc` , the income variable, `r table(is.na(brfss_17$inc))[2]` people in the BRFSS, or `r 100* (table(is.na(brfss_17$inc))[2]/length(brfss_17$inc))`% of the sample. 

The lowest number of missings is in the bad health variable, which only has `r 100* (table(is.na(brfss_17$badhealth))[2]/length(brfss_17$badhealth))`% missing.

### Mean imputation
Now, i'm going to illustrate mean imputation of a continuous variable, BMI.

```{r}
#I'm going to play with 3 outcomes, bmi, having a regular doctor and income category
summary(brfss_17$bmi) 

#what happens when we replace the missings with the mean?
brfss_17$bmi.imp.mean<-ifelse(is.na(brfss_17$bmi)==T, mean(brfss_17$bmi, na.rm=T), brfss_17$bmi)

mean(brfss_17$bmi, na.rm=T)
mean(brfss_17$bmi.imp.mean) #no difference!

fit<-lm(bmi~inc+agec+educ+race_eth, brfss_17)

```

```{r}
median(brfss_17$bmi, na.rm=T)
median(brfss_17$bmi.imp.mean) #slight difference
```

```{r}
var(brfss_17$bmi, na.rm=T)
var(brfss_17$bmi.imp.mean) # more noticeable difference!

```

So what we see here, is that imputing with the mean does nothing to central tendency (when measured using the mean, but does affect the median slightly), but it does reduce the variance in the outcome. This is because you're replacing all missing cases with the most likely value (the mean), so you're artificially deflating the variance. That's not good.

We can see this in a histogram, where the imputed values increase the peak in the distribution:

```{r}
#plot the histogram
library(reshape2)

brfss_17%>%
  select(bmi.imp.mean, bmi)%>%
  melt()%>%
  ggplot()+geom_freqpoly(aes(x = value,
     y = ..density.., colour = variable))


```



### Modal imputation for categorical data
If we have a categorical variable, an easy way to impute the values is to use modal imputation, or impute cases with the mode, or most common value. It doesn't make sense to use the mean, because what would that mean for a categorical variable?

```{r}
table(brfss_17$employ)
#find the most common value
mcv.employ<-factor(names(which.max(table(brfss_17$employ))), levels=levels(brfss_17$employ))
mcv.employ
#impute the cases
brfss_17$employ.imp<-as.factor(ifelse(is.na(brfss_17$employ)==T, mcv.employ, brfss_17$employ))
levels(brfss_17$employ.imp)<-levels(brfss_17$employ)

prop.table(table(brfss_17$employ))
prop.table(table(brfss_17$employ.imp))

barplot(prop.table(table(brfss_17$employ)), main="Original Data", ylim=c(0, .6))
barplot(prop.table(table(brfss_17$employ.imp)), main="Imputed Data",ylim=c(0, .6))
```

Which doesn't look like much of a difference because only `r table(is.na(brfss_17$employ))[2]` people were missing. Now let's try modal imputation on income group:

```{r}
table(brfss_17$inc)
#find the most common value
mcv.inc<-factor(names(which.max(table(brfss_17$inc))), levels = levels(brfss_17$inc))
mcv.inc
#impute the cases
brfss_17$inc.imp<-as.factor(ifelse(is.na(brfss_17$inc)==T, mcv.inc, brfss_17$inc))
levels(brfss_17$inc.imp)<-levels(as.factor(brfss_17$inc))

prop.table(table(brfss_17$inc))
prop.table(table(brfss_17$inc.imp))

barplot(prop.table(table(brfss_17$inc)), main="Original Data", ylim=c(0, .6))
barplot(prop.table(table(brfss_17$inc.imp)), main="Imputed Data", ylim=c(0, .6))
```

Which shows how dramatically we alter the distribution of the variable by imputing at the mode.

## Testing for MAR
### Flag variables
We can construct a flag variable. This is a useful exercise to see whether we have missing at random within the data:

```{r}
fit1<-lm(bmi~agec+educ+race_eth+is.na(inc), data=brfss_17)
summary(fit1)

```

And indeed we see that those with missing incomes have significantly lower bmi's. This implies that bmi may not be missing at random with respect to income. This is a good process to go through when you are analyzing if your data are missing at random or not.



### Multiple Imputation
These days, these types of imputation have been far surpassed by more complete methods that are based upon regression methods. These methods are generally referred to as multiple imputation, because we are really interested in imputing multiple variables simultaneously. Instead of reviewing this perspective here, I suggest you have a look at Joe Schafer's [site](http://sites.stat.psu.edu/~jls/mifaq.html) that gives a nice treatment of the subject. Here, I will use the imputation techniques in the `mice` library in R, which you can read about [here](http://www.jstatsoft.org/v45/i03/paper).

I have used these in practice in publications and generally like the framework the library uses. Another popular technique is in the `Amelia` library of [Gary King](http://gking.harvard.edu/amelia), which I haven't used much. If you are serious about doing multiple imputation it would be advised to investigate multiple methodologies.

To begin, I explore the various patterns of missingness in the data. The `md.pattern` function in `mice` does this nicely. Here, each row corresponds to a particular pattern of missingness (1 = observed, 0=missing)
```{r}
#look at the patterns of missingness
md.pattern(brfss_17[,c("bmi", "inc", "agec","educ","race_eth")])
```

The first row shows the number of observations in the data that are complete (first row). 

The second row shows the number of people who are missing *only* the inc variable. 

Rows that have multiple 0's in the columns indicate missing data patterns where multiple variables are missing. 


The bottom row tells how many total people are missing each variable, in *ANY* combination with other variables.

If you want to see how pairs of variables are missing together, the `md.pairs()` function will show this.

A pair of variables can have exactly four missingness patterns: 
Both variables are observed (pattern `rr`), the first variable is observed and the second variable is missing (pattern `rm`), the first variable is missing and the second variable is observed (pattern `mr`), and both are missing (pattern `mm`).
```{r}
md.pairs(brfss_17[,c("bmi", "inc", "agec","educ","race_eth")])
```

### Basic imputation:
We can perform a basic multiple imputation by simply doing: **Note this may take a very long time with big data sets**

```{r}

dat2<-brfss_17
samp2<-sample(1:dim(dat2)[1], replace = F, size = 500)
dat2$bmiknock<-dat2$bmi
dat2$bmiknock[samp2]<-NA

head(dat2[, c("bmiknock","bmi")])
imp<-mice(data = dat2[,c("bmi", "inc", "agec","educ","race_eth")], seed = 22, m = 5)

print(imp)
```

Shows how many imputations were done. It also shows total missingness, which imputation method was used for each variable (because you wouldn't want to use a normal distribution for a categorical variable!!).

It also shows the sequence of how each variable is visited (or imputed, the default is left to right). 

We may want to make sure imputed values are plausible by having a look. For instance, are the BMI values outside of the range of the data.

```{r}
head(imp$imp$bmi)
summary(imp$imp$bmi)
summary(brfss_17$bmi)
```

```{r}
head(imp$imp$inc)
summary(imp$imp$inc)
```

Which shows the imputed values for the first 6 cases across the 5 different imputations, as well as the numeric summary of the imputed values. We can see that there is variation across the imputations, because the imputed values are not the same.


We can also do some plotting. For instance if we want to see how the observed and imputed values of bmi look with respect to race, we can do:
```{r, fig.height=7, fig.width=8}
library(lattice)
stripplot(imp,bmi~race_eth|.imp, pch=20)
```

and we see the distribution of the original data (blue dots), the imputed data (red dots) across the levels of race, for each of the five different imputation runs(the number at the top shows which run, and the first plot is the original data).

This plot shows that the bmi values correspond well with the observed data, *so they are probably plausible values*.

If we want to get our new, imputed data, we can use the `complete()` function, which by default extracts the first imputed data set. If we want a different one, we can do `complete(imp, action=3)` for example, to get the third imputed data set.

```{r}
dat.imp<-complete(imp, action = 1)
head(dat.imp, n=10)

#Compare to the original data
head(brfss_17[,c("bmi", "inc", "agec","educ","race_eth")], n=10)
```

While the first few cases don't show much missingness, we can coax some more interesting cases out and compare the original data to the imputed:

```{r}
head(dat.imp[is.na(brfss_17$bmi)==T,], n=10)
head(brfss_17[is.na(brfss_17$bmi)==T,c("bmi", "inc", "agec","educ","race_eth")], n=10)

```



### Analyzing the imputed data
A key element of using imputed data, is that the relationships we want to know about should be maintained after imputation, and presumably, the relationships within each imputed data set will be the same. So if we used each of the (5 in this case) imputed data sets in a model, then we should see similar results across the five different models.

Here I look at a linear model for bmi:
```{r}
#Now, I will see the variability in the 5 different imputations for each outcom
fit.bmi<-with(data=imp ,expr=lm(bmi~inc+agec+educ+race_eth))
fit.bmi
```

### variation in bmi
```{r}

with (data=imp, exp=(sd(bmi)))
```

### Frequency table for income
```{r}
with (data=imp, exp=(prop.table(table(inc))))
```

### Frequency table for race/ethnicty

```{r}
with (data=imp, exp=(prop.table(table(race_eth))))
```

### Frequency table for education

```{r}
with (data=imp, exp=(prop.table(table(educ))))

```


Now we pool the separate models from each imputed data set:
```{r}
est.p<-pool(fit.bmi)
print(est.p)
summary(est.p)
```

We need to pay attention to the `fmi` column and the `lambda` column. These convey information about how much the missingness of each particular variable affects the model coefficients. 

```{r}
lam<-data.frame(lam=est.p$pooled$lambda, param=row.names(est.p$pooled))

ggplot(data=lam,aes(x=param, y=lam))+geom_col()+theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

It appears that a couple of the education variables and the income variables have large variances to them. This suggests that there may be noticeable variation in the resulting coefficient of the model, depending on which imputed data set we use.

We can also compare to the model fit on the original data, with missings eliminated:
```{r}
library(dplyr)
bnm<-brfss_17%>%
  select(bmi, inc, agec, educ, race_eth)%>%
  filter(complete.cases(.))%>%
  as.data.frame()

summary(lm(bmi~inc+agec+educ+race_eth, bnm))
```

### Compare imputed model to original data
Here, I compare the coefficients from the model where we eliminated all missing data to the one that we fit on the imputed data:
```{r}
fit1<-lm(bmi~inc+agec+educ+race_eth, data=brfss_17)
summary(fit1)

fit.imp<-lm(bmi~inc+agec+educ+race_eth, data=dat.imp)
summary(fit.imp)
```

In the analysis that only uses complete  cases, we see a significant income effect on bmi, but not once we impute the missing values. This suggests a significant selection effect for the income variable.



## Examining the variation in the models for the imputed data
If we wanted to see the ranges of the betas in the five imputed data models, we could do that:
```{r, fig.height=6, fig.width=9}
#get the coefficients from each of the 5 imputations of bmi
coefs<-as.data.frame(matrix(unlist(lapply(fit.bmi$analyses, coef)), nrow=5, ncol=17, byrow=T))
names(coefs)<-names(fit.bmi$analyses[[1]]$coef)
#plot the coefficients from each of the different rounds of imputation to see the variability in the
#results

coefs%>%
  select(`inc2_15-25k`:`race_ethnh other`)%>%
  melt()%>%
  mutate(imp=rep(1:5, 16))%>%
  ggplot()+geom_point(aes(y=value,x=variable, group=variable, color=as.factor(imp) ))+ggtitle("Estimated Betas from Each Imputed Regression Model for BMI Outcome")+theme(axis.text.x = element_text(angle = 45, hjust = 1))




```


