---
title: 'DEM 7283: Longitudinal Models for Change using Generalized Estimating Equations'
author: "Corey Sparks, PhD"
date: "April 27, 2020"
output:
   html_document: 
    df_print: paged
    fig_height: 7
    fig_width: 7
    toc: yes
    includes:
      in_header: logo.html
---

In this example, we will use Generalized Estimating Equations to do some longitudinal modeling of data from the [ECLS-K 2011](https://nces.ed.gov/ecls/dataproducts.asp). Specifically, we will model changes in a student's standardized math score as a continuous outcome and self rated health as a binomial outcome, from fall kindergarten to spring, 1st grade. 

Introduction to GEE's
========================================================

Up until now, we have used (G)LMM's to analyze data that were "clustered"

- Persons within neighborhoods
- Survey data in general - stratified sampling

The next topic will introduce a modeling strategy that allows us to consider clustered data, but in a different fashion

GLMMS's
========================================================

GLMM's are commonly referred to as *conditional* models, because the model coefficients "$\beta$'s" are condition on the random effects in the model. 

Likewise, the mean if conditional on the random effects. This is another way of saying that the mean for a given covariate pattern is conditional on the group that the particular person is in. 

$\mu_{ij}^c = E(Y_{ij} | u_j) = X_{ij}\beta + u_j$

GLMMS's and GEE's
========================================================

In contrast, **Generalzed Estimating Equations** are referred to as *marginal* models because they only estimate the overall mean.

$\mu_{ij} = X_{ij}\beta$

[Lee and Nelder, 2004](http://www.people.vcu.edu/~dbandyop/BIOS625/LeeNelder2004.pdf) provide a very good description of how these two methods compare to one another

Generalized Estimating Equations
========================================================

- Typically first attributed to [Liang and Zeger, 1986](http://biomet.oxfordjournals.org/content/73/1/13.short)
- GEE's are regression models
- Interested in modeling the *mean response*, while treating correlation within person/cluster as a *nuisance*
- **NOT** based on maximum likelihood
- Does not need a fully specified joint distribution, only the marginal distribution (*mean*) of the outcome
- Models can be for any distribution for the outcome

GEE's
========================================================

- For longitudinal data, we assume we have $y_{ij}$ as our outcome on person *i* at time *j*. This could just as easily be persons within other types of clusters, like counties or sampling units.
- Also have $X_{ij}$, the matrix of predictors
- Specify the link function between $y_{ij}$ and $X_{ij}$ as in a GLM, via a *link function*
- Focus is on the *linear predictor* of the link function - the mean
- **NOT INTERESTED** in variance components **ONLY** regression coefficients

GEE's
========================================================

- Covariance structure
  - We also may wish to model *how* observations are related to one another via some type of *correlation structure* between waves
  - This directly implies that observations are *NOT INDEPENDENT*, and that's fine
  - Observations between clusters **are** independent
  - Errors are correlated
  - No assumption of common variance (homoskedsasticity)


GEE's - Model form
========================================================

A basic form of the model would be:

$Y_{ij} = \beta_0 + \sum_k X_{ijk} \beta_k + CORR + error$

Ordinary models will tend to over estimate the standard errors for the $\beta$'s  for time varying predictors in a model with repeated observations, because these models do not account for the correlation within clusters \ observations over time. 

Likewise, the standard errors of time invariant predictors will be under estimated


GEE's - Model estimation
========================================================

Given the mean function for the model and a specified correlation function, the model parameters may be estimated by finding the solution for:

$$U(\beta) = \sum_i ^n   \frac{\delta \mu_{ij}}{ \delta \beta_k} V_i^{-1} (Y_{ij} - \mu(\beta))$$


Which gives estimates of the $\beta$'s for the linear mean function. 

GEE's - Model estimation
========================================================

- First, a naive linear regression analysis is carried out, assuming the observations within subjects are independent.
- Then, residuals are calculated from the naive model (observed-predicted) and a working correlation matrix is estimated from these residuals.
- Then the regression coefficients are refit, correcting for the correlation. (Iterative process)
- The within-subject correlation structure is treated as a nuisance variable (i.e. as a covariate)  

GEE's - Correlation Structure
========================================================

For three time points per person, the ordinary regression model correlation in residuals within clusters/persons over time can be thought of as the matrix:

$$\begin{bmatrix}
\sigma^2 & 0 & 0 \\ 
0 & \sigma^2 &0 \\ 
 0 & 0 & \sigma^2
\end{bmatrix}$$

which assumed the variances are constant and the residuals are independent over time

GEE's - Correlation Structure
========================================================

But in a GEE, the model include the actual correlation between measurements over time: 

$$\begin{bmatrix}
\sigma_1 ^2 & a & c \\ 
a & \sigma_2 ^2 &b \\ 
 b & c & \sigma_3 ^2
\end{bmatrix}$$

Which allows the variances over time to be different, as well as correlations between times to be present.


GEE's - Correlation Structure
========================================================

- Several types of correlation/covariance are commonly used in GEE's
- When we fit a GEE, we have to assume a certain type of correlation for the repeated measures. These are typically:
  - Independence - same as OLS
  - Exchangeable/compound symmetry (simplest)
  - Autoregressive
  - Unstructured (most complicated)
  
GEE's - Correlation Structure - Independent
========================================================

$$\begin{bmatrix}
1 & 0 & 0 \\ 
0 & 1 &0 \\ 
 0 & 0 & 1
\end{bmatrix}$$


GEE's - Correlation Structure - Exchangeable
========================================================

$$\begin{bmatrix}
1 & \rho & \rho \\ 
\rho & 1 &\rho \\ 
 \rho &\rho & 1
\end{bmatrix}$$


GEE's - Correlation Structure - AR(1)
========================================================

$$\begin{bmatrix}
1 & \rho & \rho^2 \\ 
\rho & 1 &\rho\\ 
 \rho^2 & \rho & 1
\end{bmatrix}$$

GEE's - Correlation Structure - Unstructured
========================================================

$$\begin{bmatrix}
1 & \rho_1 & \rho_2 \\ 
\rho_1 & 1 &\rho_3 \\ 
 \rho_2 & \rho_3& 1
\end{bmatrix}$$

```{r}

library (car)
library(geepack)
library(MuMIn)  #need to install
library(dplyr)

```


## Data and recodes
First we load our data
```{r, message=FALSE}
load("~/OneDrive - University of Texas at San Antonio//classes/dem7283/class_20_7283//data/eclsk_k5.Rdata")
names(eclskk5)<-tolower(names(eclskk5))
#get out only the variables I'm going to use for this example

#subset the data

eclsk.sub<-eclskk5%>%
  select(childid, x_chsex_r, x1locale, x_raceth_r, x2povty, x12par1ed_i, p1curmar, x1htotal, x1mscalk5, x2mscalk5, x3mscalk5, x4mscalk5, x5mscalk5, p1hscale, p2hscale, p4hscale, x2fsstat2, x4fsstat2, x4fsstat2, x12sesl, x4sesl_i, p2parct1, p2parct2, s1_id, p2safepl, x2krceth, p1o2near, x_distpov, w1c0, w1p0, w2p0, w1c0str, w1p0str, w4c4p_40, w4c4p_4str,w4c4p_4psu, w1c0psu, w1p0psu, x1height, x2height, x4height, x4height, x5height, x1kage_r, x2kage_r, x3age, x4age, x5age)

gc()
#rm(eclsk11); gc()
```


 
## Time constant variables

First, I do some recoding of variables. First, we code time **invariant** variables, meaning their values do not change at each wave. 

```{r}
#Non time varying variables
#First we recode some Child characteristics
#Child's sex: recode as male =1
eclsk.sub$male<-Recode(eclsk.sub$x_chsex_r, recodes="1=1; 2=0; -9=NA")

#Recode race with white, non Hispanic as reference using dummy vars
eclsk.sub$hisp<-Recode (eclsk.sub$x_raceth_r, recodes="3:4=1;-9=NA; else=0")
eclsk.sub$black<-Recode (eclsk.sub$x_raceth_r, recodes="2=1;-9=NA; else=0")
eclsk.sub$asian<-Recode (eclsk.sub$x_raceth_r, recodes="5=1;-9=NA; else=0")
eclsk.sub$nahn<-Recode (eclsk.sub$x_raceth_r, recodes="6:7=1;-9=NA; else=0")
eclsk.sub$other<-Recode (eclsk.sub$x_raceth_r, recodes="8=1;-9=NA; else=0")


#Then we recode some parent/mother characteristics
#Mother's education, recode as 2 dummys with HS = reference
eclsk.sub$lths<-Recode(eclsk.sub$x12par1ed_i, recodes = "0:2=1; 3:8=0; else = NA")
eclsk.sub$gths<-Recode(eclsk.sub$x12par1ed_i, recodes = "1:3=0; 4:8=1; else =NA") 

#marital status, recode as 2 dummys, ref= married
eclsk.sub$single<-Recode(eclsk.sub$p1curmar, recodes="4=1; -7:-9=NA; else=0")
eclsk.sub$notmar<-Recode(eclsk.sub$p1curmar, recodes="2:3=1; -7:-9=NA; else=0")


#Then we do some household level variables

#Urban school location = 1
eclsk.sub$urban<-Recode(eclsk.sub$x1locale, recodes = "1:3=1; 4=0; -1:-9=NA")

#poverty level in poverty = 1
eclsk.sub$pov<-Recode(eclsk.sub$x2povty , recodes ="1:2=1; 3=0; -9=NA")

#Household size
eclsk.sub$hhsize<-eclsk.sub$x1htotal

#school % minority student body
eclsk.sub$minorsch<-ifelse(eclsk.sub$x2krceth <0, NA, eclsk.sub$x2krceth/10)

#Unsafe neighborhood
eclsk.sub$unsafe<-Recode(eclsk.sub$p2safepl , recodes = "1:2='unsafe'; 3='safe'; else=NA",as.factor = T)

#school district poverty
eclsk.sub$dist_pov<-ifelse(eclsk.sub$x_distpov==-9, NA, scale(eclsk.sub$x_distpov))
```

## Time varying variables
I have to make the repeated measures of each of my longitudinal variables. These are referred to as **time varying** variables, meaning their values change at each wave.

```{r}
#Longitudinal variables
#recode our outcomes, the  first is the child's math standardized test score  in Kindergarten
eclsk.sub$math1<-ifelse(eclsk.sub$x1mscalk5<0, NA, eclsk.sub$x1mscalk5)
eclsk.sub$math2<-ifelse(eclsk.sub$x2mscalk5<0, NA, eclsk.sub$x2mscalk5)
#eclsk.sub$math3<-ifelse(eclsk.sub$x3mscalk1<0, NA, eclsk.sub$x3mscalk1)
eclsk.sub$math4<-ifelse(eclsk.sub$x4mscalk5<0, NA, eclsk.sub$x4mscalk5)

#Second outcome is child's height for age, continuous outcome
eclsk.sub$height1<-ifelse(eclsk.sub$x1height<=-7, NA, eclsk.sub$x1height)
eclsk.sub$height2<-ifelse(eclsk.sub$x2height<=-7, NA, eclsk.sub$x2height)
#eclsk.sub$height3<-ifelse(eclsk.sub$x3height<=-7, NA, eclsk.sub$x3height)
eclsk.sub$height4<-ifelse(eclsk.sub$x4height<=-7, NA, eclsk.sub$x4height)

#Age at each wave
eclsk.sub$age_yrs1<-ifelse(eclsk.sub$x1kage_r<0, NA, eclsk.sub$x1kage_r/12)
eclsk.sub$age_yrs2<-ifelse(eclsk.sub$x2kage_r<0, NA, eclsk.sub$x2kage_r/12)
#eclsk.sub$age_yrs3<-ifelse(eclsk.sub$x3age<0, NA, eclsk.sub$x3age/12)
eclsk.sub$age_yrs4<-ifelse(eclsk.sub$x4age<0, NA, eclsk.sub$x4age/12)

eclsk.sub<- eclsk.sub[is.na(eclsk.sub$age_yrs1)==F, ]

#Height for age z score standardized by sex and age
eclsk.sub$height_z1<-ave(eclsk.sub$height1, as.factor(paste(round(eclsk.sub$age_yrs1, 1.5), eclsk.sub$male)), FUN=scale)
eclsk.sub$height_z2<-ave(eclsk.sub$height2, as.factor(paste(round(eclsk.sub$age_yrs2, 1.5), eclsk.sub$male)), FUN=scale)
#eclsk.sub$height_z3<-ave(eclsk.sub$height3, as.factor(paste(round(eclsk.sub$age_yrs3, 1.5), eclsk.sub$male)), FUN=scale)
eclsk.sub$height_z4<-ave(eclsk.sub$height4, as.factor(paste(round(eclsk.sub$age_yrs4, 1.5), eclsk.sub$male)), FUN=scale)


#Household food insecurity, dichotomous outcome
#This outcome is only present at two waves
eclsk.sub$foodinsec1<-Recode(eclsk.sub$x2fsstat2, recodes="2:3=1; 1=0; else=NA")
eclsk.sub$foodinsec2<-Recode(eclsk.sub$x4fsstat2, recodes="2:3=1; 1=0; else=NA")


#Child health assessment Excellent to poor , ordinal outcome
eclsk.sub$chhealth1<-ifelse(eclsk.sub$p1hscale<0, NA, eclsk.sub$p1hscale)
eclsk.sub$chhealth2<-ifelse(eclsk.sub$p2hscale<0, NA, eclsk.sub$p2hscale)
eclsk.sub$chhealth4<-ifelse(eclsk.sub$p4hscale<0, NA, eclsk.sub$p4hscale)

#SES
eclsk.sub$hhses1<-ifelse(eclsk.sub$x12sesl==-9, NA, scale(eclsk.sub$x12sesl))
eclsk.sub$hhses2<-ifelse(eclsk.sub$x12sesl==-9, NA, scale(eclsk.sub$x12sesl))

eclsk.sub$hhses4<-ifelse(eclsk.sub$x4sesl_i==-9, NA, scale(eclsk.sub$x4sesl_i))
```

## Reshaping data into longitudinal format
To analyze data longitudinally, we must reshape the data from its current "wide" format, where each repeated measure is a column, into the "long" format, where there is a single column for a particular variable, and we account for the repeated measurements of each person. In this case, I'm going to use three waves of data, so each child can contribute up to three lines to the data.

The `reshape()` function will do this for us, but below I use a tidy method, using a combination of the `data.table` and `dplyr` packages. I first make a long data set of the height, age, math, child health and household SES measures, then I left join it to the time invariant variables i'll use in my models below. 
```{r}

library(data.table)
library(magrittr)
out<-melt(setDT(eclsk.sub), id = "childid",
          measure.vars = list(ht=c("height_z1","height_z2","height_z4"),
                              age=c("age_yrs1", "age_yrs2", "age_yrs4"), 
                              math=c("math1", "math2", "math4"),
                              hhses=c("hhses1", "hhses2", "hhses4"),
                              health=c("chhealth1", "chhealth2", "chhealth4")))%>%
  setorder(childid)
head(out, n=20)

e.long<-eclsk.sub%>%
  select(childid, hisp, black,asian, nahn, other,male, unsafe, s1_id, pov, hhsize, urban, w4c4p_40, w4c4p_4str, w4c4p_4psu)%>%
  left_join(., out, "childid")

e.long$wave<-rep(c(1,2,4), length(unique(e.long$childid)))
head(e.long)


head(e.long)

e.long.comp<-e.long%>%
  filter(complete.cases(.), w4c4p_40>0)


```


### Visualization of longitudinal data
```{r}
library(ggplot2)

first10<-unique(e.long.comp$childid)[1:10]

sub<-e.long.comp%>%
  filter(childid%in%first10)

ggplot(sub, aes(x=age, y=math))+geom_point()+ geom_smooth(method='lm',formula=y~x)+facet_wrap(~childid,nrow = 3)+ggtitle(label = "Change in Math score across age", subtitle = "First 10 children in ECLS-K 2011")

```

## Modeling

### Longitudinal model using GLMM
```{r}
library(lme4)

library(brms)
e.long.comp$wt<-e.long.comp$w4c4p_40/mean(e.long.comp$w4c4p_40, na.rm=T)
modb<-brm(scale(math)~scale(age)+male+black+hisp+asian+nahn+other+hhses+(1|childid),family = "gaussian",cores = 3,chains = 3,data=e.long.comp)
#random intercept model for each person
mod1<- lmer(scale(math)~scale(age)+male+black+hisp+asian+nahn+other+hhses+(1|childid), data=e.long.comp, weights=w4c4p_40/mean(w4c4p_40), REML = F)

summary(mod1)


#growth curve model
mod2<- lmer(scale(math)~scale(age)+male+black+hisp+asian+nahn+other+hhses+(scale(age)|childid), data=e.long.comp, weights=w4c4p_40/mean(w4c4p_40), REML = F)

summary(mod2)


#multi-level model with child within school - this can be used for people within states or neighborhoods
mod3<- lmer(scale(math)~scale(age)+male+black+hisp+asian+nahn+other+hhses+(1|childid)+(1|s1_id), data=e.long.comp, weights=w4c4p_40/mean(w4c4p_40))

summary(mod3)

library(lmerTest)
anova(mod1, mod2, mod3, ddf="Kenward-Roger")
```


### Longitudinal Models using GEE's

Again, the GEE is used here instead of the (G)LMM. 
```{r, echo=FALSE, results='hide'}

#basic linear model
fit.1<-glm(scale(math)~scale(age)+male+black+hisp+asian+nahn+other+hhses, data=e.long.comp, weights=w4c4p_40/mean(w4c4p_40))
summary(fit.1)

#Get residuals and put them in a data frame
e.long.comp$resid<- residuals(fit.1)

test<-reshape(e.long.comp, idvar = "childid", v.names = "resid", timevar = "wave", direction = "wide", drop = names(e.long.comp)[2:21])
head(test[order(test$childid),], n=12)
```

Here is our actual correlation matrix in the residuals between waves: 
```{r}
cor(test[,-1], use="pairwise.complete")
```

This is certainly not independence, and looks more like an AR(1), because the correlation decreases as the *difference* between wave number increases.

Now we fit the GEE:
### Model with independent correlation
Meaning ZERO correlation between waves
```{r}

fit.1<-geeglm(scale(math)~scale(age)+male+black+hisp+asian+nahn+other+hhses,id=childid , wave = wave, corstr ="independence",   data=e.long.comp, weights=w4c4p_40/mean(w4c4p_40))
summary(fit.1)
```


### Model with Exchangeable correlation
Meaning correlation between waves, but the correlation is the same for each pair waves 
```{r}

fit.2<-geeglm(scale(math)~scale(age)+male+black+hisp+asian+nahn+other+hhses,id=childid , wave = wave, corstr ="exchangeable",  data=e.long.comp, weights=w4c4p_40/mean(w4c4p_40))
  
summary(fit.2)

```

The second model shows the exchangeable correlation to be `r fit.2$geese$alpha`, which is not very different from our measured correlations above 
`r knitr::kable(cor(test[, -1], use="pairwise.complete"))`

Now we examine the AR1 correlation types:
```{r}
fit.3<-geeglm(scale(math)~scale(age)+male+black+hisp+asian+nahn+other+hhses, id=childid , wave = wave, corstr ="ar1",  data=e.long.comp, weights=w4c4p_40/mean(w4c4p_40))
  
summary(fit.3)

```

The implied correlation in the AR(1) model is : `r fit.3$geese$alpha`

The other type of correlation allowed in `geeglm` is the unstructured correlation model. This is how to fit the unstructured model, but it crashes on this data, so beware
```{r, eval=FALSE}
fit.4<-geeglm(scale(math)~scale(age)+male+black+hisp+asian+nahn+other+ses, id=childid , wave = wave, corstr ="unstructured",  data=e.long.comp, weights=w4c4p_40/mean(w4c4p_40))
summary(fit.4)

```

Since GEE's aren't fit via maximum likelihood, they aren't comparable in terms of AIC or likelihood ratio tests. However, [Pan, 2001](http://onlinelibrary.wiley.com/doi/10.1111/j.0006-341X.2001.00120.x/abstract) describe an information criterion using a Quasi-likelihood formulation. This can be used to compare models with alternative correlation structures, with the lowest **QIC** representing the best fitting model. Another criterion is the **Correlation Information Criterion** (Hin and Wang, 2008)[https://onlinelibrary.wiley.com/doi/abs/10.1002/sim.3489], which is proposed to be better for choosing among models with the same mean function, but different correlation structures, which is what we're doing here.

```{r}
library(MESS) #need to install
QIC(fit.1)
QIC(fit.2)
QIC(fit.3)
```
So, it looks like the exhangeable correlation structure is slightly  better than the AR(1), using the CIC but there is not much difference between models using this criteria. 


### Binary response longitudinal model

Here we use the GEE for a binomial outcome. 

Here are what the data look like:
```{r}

binomial_smooth <- function(...) {
  geom_smooth(method = "glm", method.args = list(family = "binomial"), ...)
}

sub$poorhealth<-Recode(sub$health, recodes="2:3=1; else=0")

ggplot(sub, aes(x=age, y=poorhealth))+geom_point()+ binomial_smooth()+ggtitle(label = "Change in Math score across age", subtitle = "First 10 children in ECLS-K 2011 - All children")

ggplot(sub, aes(x=age, y=poorhealth))+geom_point()+ binomial_smooth()+facet_wrap(~childid,nrow=3)+ggtitle(label = "Change in Math score across age", subtitle = "First 10 children in ECLS-K 2011 - Invidivual Children")

```


```{r}
btest<-glm(I(health>2)~age+male+black+hisp+asian+nahn+other+hhses+factor(wave) , family=binomial, data=e.long.comp, weights=w4c4p_40/mean(w4c4p_40))

e.long.comp$binomresid<- residuals(btest)
test2<-reshape(e.long.comp, idvar = "childid", v.names = "binomresid", timevar = "wave", direction = "wide", drop = names(e.long.comp)[c(2:21, 23)])

head(test2[order(test$childid),], n=12)

#empirical correlations between waves in the residuals
cor(test2[, -1], use="pairwise.complete", method = "spearman")


```
These look like a constant correlation, or AR(1)

### Logistic GEE with independent correlation
```{r}

fitb.1<-geeglm(I(health>2)~age+male+black+hisp+asian+nahn+other+hhses, waves = wave,id=childid ,corstr ="independence", family=binomial, data=e.long.comp, weights=w4c4p_40/mean(w4c4p_40))
summary(fitb.1)


```

### Logistic GEE with exchangeable correlations
```{r}

fitb.2<-geeglm(I(health>2)~age+male+black+hisp+asian+nahn+other+hhses, waves = wave,id=childid, family=binomial, data=e.long.comp, corstr="exchangeable", weights=w4c4p_40/mean(w4c4p_40))
summary(fitb.2)


```

### Logistic GEE with AR(1) correlation
```{r}
fitb.3<-geeglm(I(health>2)~age+male+black+hisp+asian+nahn+other+hhses, waves = wave,id=childid, family=binomial, data=e.long.comp, corstr="ar1", weights=w4c4p_40/mean(w4c4p_40))
summary(fitb.3)
```

Compare the three models:
```{r}
QIC(fitb.1)
QIC(fitb.2)
QIC(fitb.3)
```

In the binomial case, it looks like the echangeable correlation structure is good, as it has the lowest CIC, although the AR(1) model is very similar.

